Evaluating Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe using radiocarbon data

### 2019-03-28 review/answer_to_review.md

> Firstly, my impression is that the author is trying to squeeze perhaps a bit too much in a single paper. I felt this was particularly the case of the computer simulation model which is almost thrown in in the mix, without much discussion on model parameters (see below). I would be tempted to suggest removing this entirely and include this into another paper.

After careful deliberation, I have decided to follow this advice. The simulation (1) does not add information that would be absolutely essential for the core topics, (2) reduces the overall comprehensibility due to the terminology that has to be introduced to describe it and (3) cannot be adequately presented and discussed here without leaving the initial scope of the paper. I will prepare another paper in the future that will present this simulation and other simulation approaches.

> Secondly, I feel like the empirical analyses have too much preprocessing. The raw data provides spatial and temporal coordinates of individual burials, yet these are combined spatially and chronologically first and analysed as aggregates. As a consequence: 1) there is a substantial loss of spatial information; 2) sampling error is effectively disregarded (i.e. mantel tests are based on proportion values that are calculated for regions with very different samples sizes) and strictly speaking the samples are also non-independent (i.e. a large site with multiple burials can strongly bias the proportions of a particular region); and 3) there is an increased risk of ecological fallacy/Simpson’s paradox. I take this aggregation provides a direct mapping to existing literature, but I am not convinced that this is worth sacrificing so much.

The decision to create discrete spatial area classes is necessary to establish subsets of dates that can be used for time series calculation. I'm not aware of a methodologival approach where time series could be constructed without some sort of aggregation. The equal area size approach applied here for the region defition causes size of the areas to be very big due to the currently overall small amount of data. I expect this to change in the future and the analysis can be redone then with more samples and much smaller areas to understand the definitely much more complex processes within the macro regions established here.

So: The criticism listed here is valid and I added it in the discussion. I considered a complete revision of the time series construction methodology, but since I (1) am still convinced that it is scientifically sound and reproducible, and (2) do not know a fundamentally better approach that avoids these completely risks, I suggest publishing the results as they are for the community of practice to discuss and improve.

@@

> Page 3, Column 2, Lines 40-44. The definition of these regions seems rather arbitrary and necessary. Why create such an artificial unit when the spatial coordinates are available?

Please see the reply above.

> Page 3, Column 2, Line 60. “For the remaining 154 dates...”. I assume these dates were excluded as we don’t know whether they are terrestrial or marine?  Also, the paragraph mentions “bones and teeth”, are there any risks of aquatic diet determining a reservoir effect?

These dates were not excluded to avoid further reduction of the amount of data. It is not possible to consider these dating effects on this scale, as it would require to study all radiocarbon dates and their archaeological context individually. I agree that it might be necessary to do so, but this is not an endeavour that can be undertaken without long-term project funding. Marine reservoir effect is only one of the possible problems relevant here in addition to the Old wood effect or the human 14C-residence ("adult") effect. Not to mention cultural phenomena like secondary treatment of the body. I added a paragraph about this in the *Data quality* section.

### 2019-03-28 article/06_Supplementary_Remarks.Rmd
dating problems

Concerning the primary variables *burial type* and *burial construction*, further special effects must be taken into account: The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be under-represented in Radon-B. In many places, they were also the subject of expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an over-representation of mounds seems more likely than an over-representation of flat graves. The dating of cremation graves on bone remains methodically differs from the dating of inhumations because it is based on the extraction of the mineral, not the organic, fraction in the bone [@lantingDatingCrematedBones2001]. The underlying chemical method was developed later, has only been used more frequently in the past 15 years and is associated with some particular difficulties [@olsenCharacterisationBlindTesting2008; @strydonck14CDatingCremated2009]. Burnt bone is generally more susceptible to weathering than unburned bone. An under-representation of cremation graves in current data collections is, therefore, more plausible.

>>>>

Concerning the primary variables *burial type* and *burial construction*, further special effects must be taken into account:

The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be under-represented in Radon-B. In many places, they were also the subject of expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an over-representation of mounds seems more likely than an over-representation of flat graves.

The dating of cremation graves on bone remains methodically differs from the dating of inhumations because it is based on the extraction of the mineral, not the organic, fraction in the bone [@lantingDatingCrematedBones2001]. The underlying chemical method was developed later, has only been used more frequently in the past 15 years and is associated with some particular difficulties [@olsenCharacterisationBlindTesting2008; @strydonck14CDatingCremated2009]. Burnt bone is generally more susceptible to weathering than unburned bone. An under-representation of cremation graves in current data collections is, therefore, more plausible.

Another aspect to consider concerning the data quality is the temporal accuracy of the radiocarbon dating. Ideally each radiocarbon date represents the exact point in time of the creation of the respective grave. Unfortunately several natural and cultural effects can cause a (partially systematic) offset between the radiocarbon date and the burial ceremony, including the *Marine reservoir effect*, the *Old wood effect*, the *Human ^14^C-residence ("adult") effect* or cultural practices like a secondary treatment of the body or the deposition of burial objects that are already ancient at the time of the ceremony. These effects can have strong influence on the accuracy of the dating information, but can only be explored for each grave individually. In the present paper they therefore had to be ignored.
